Reasoning about a Capability Machine with Local Capabilities

### 2017-05-31 popl-2018/paper.tex
An attempt at an introductory section about the logical relation.

\section{Logical relation}
\label{sec:logical-relation}
In order to reason about programs on the capability machine and prove properties
such as well-bracketedness and encapsulation, we define a unary logical
relation.
% Intuition
It is a semantic model of all the words that are definitely safe to give to an
adversary. Whether a capability is safe depends heavily on the contents of the
memory since instructions as well as capabilities may reside there. Hence we
index the logical relation over worlds, modelling the memory (this makes our
logical relation into a Kripke logical relation). A world consists of a finite
set of regions, and each region defines a protocol that governs the possible
evolution of a part of memory. A future world relation defines how worlds may
evolve during execution (e.g. new protocols established over memory received
from malloc, or steps in a protocol already established for a part of memory).
In fact, as we will explain below, there are two future world relations to
accommodate reasoning about local capabilities.
%

\subsection{Worlds}
\begin{figure}
  \centering
  \includegraphics{w11}
  \caption{An illustration of local/global capabilities and temporary/permanent
    regions. The table represents a memory and the colored fields are regions
    governing a memory range. Global capabilities cannot depend on temporary
    regions.}
  \label{fig:cap-world}
\end{figure}
% \dominique{Perhaps we could first try to convey the intuition of a permanent
%   world with a temporary overlay, with global capabilities ignoring the overlay.
%   Perhaps we could add a depiction of the world to Figure 5?}
%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}
A world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, to describe a protocol for the region's memory,
similarly to what has been used in logical relations for high-level
languages~\citep{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. Protocols
imposed by permanent regions stay in place indefinitely. Any capability, local
or global, can depend on these protocols. Protocols imposed by temporary regions
can be revoked at certain points during execution, on the condition that we
clear local capabilities when this happens. This means that local capabilities
can safely depend on the protocols imposed by temporary regions, but global
capabilities cannot, since a global capability may outlive a temporary region
that is revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

>>>>

\section{Logical relation}
\label{sec:logical-relation}
Now that we have defined our calling convention, how can we be sure that it
works? More concretely, suppose that we have a program that uses the convention
in its interaction with untrusted adversary code. Can we formally prove the
program's correctness if it relies on well-bracketed control flow and private
state encapsulation for the interaction with the adversary? Clearly, such a
proof should depend on a formal expression of the guarantees provided by the
capability machine, including the specific guarantees for local capbalities.

In this section, we construct such a formalisation. We make use of some
well-studied and powerful (but non-trivial) machinery from the literature.
Specifically, we employ a unary step-indexed Kripke logical relation with
recursive worlds, and some additional special characteristics of our own.
Step-indexing, Kripke logical relations and recursive worlds are techniques that
may be familiar from lambda calculus settings, but it may not be clear to the
reader how they apply in this more low-level assembly language. Therefore, we do
not immediately dive into the details, but we first try to provide some informal
intuition about how all of this machinery comes into play in our setting, in the
next section.

\subsection{Formalizing the guarantees of the capability machine}
What differentiates a capability machine from a more standard assembly language
is that we can bound the authority of an executing block of code, based solely
on the capabilities it has access to. Specifically, it does not matter which
instructions are actually executed, i.e. the bound also applies to untrusted
adversary code that has not been inspected or modified in any way.

\paragraph{Worlds}
But what does a ``bound on the authority'' of an executing block of code mean?
In our setting, there are no externally observable side-effects and the only
primitive authority that code may hold is authority over memory. As such, the
authority bounds we consider are related to memory, but in a form that is more
fine-grained than standard read/write authority: code's authority can be bounded
by arbitrary memory invariants that it is required to respect. Specifically, we
will define worlds $W \in \Worlds$ that define a set of memory invariants and
our results will express authority bounds on code as safety with respect to such
a world, i.e. the fact that the code respects the invariants registered in
the world.

\paragraph{Safe values}
So, let's say that we have a world $W$ expressing that the memory must contain
value $42$ at address $0$ must remain $42$, may contain arbitrary values at
address $50$-$60$, a $\readwrite$ capability for address $0$ at address $73$ and
an integer at address $100$ that may only increase over time\footnote{Indeed, we
  will allow a notion of \emph{evolvable} invariants or protocols that can
  express such a temporal property.}. Our main theorem will state that if the
current register file only contains safe words (numbers or capabilities which
preserve the invariants in $W$ under any interaction), then the execution will
necessarily also preserve the invariants (irrespective of the instructions being
executed).

To make this more precise, we need to define the set $\stdvr(W) \in
\powerset{\Words}$ of words that are safe w.r.t. $W$. Essentially, the set
should only include words that preserve $W$'s invariants under any interaction,
but should otherwise be as liberal as possible. Numbers are clearly always safe,
as they cannot be used to break invariants. Whether a capability is safe depends
on the authority that it grants. In the above-described world, a read capability
for address $0$ is safe, as it can only be used to read the value $42$, which is
itself safe. However, a write capability for address $0$ is not safe: it can be
used to overwrite the memory at that address with a value other than $42$,
breaking the invariant for that address.

\paragraph{Step-indexing}
More generally, we want to define that a read capability for memory range
$[b,e]$ is safe if the world guarantees that the words at those addresses are
themselves safe. However, this definition is cyclic: what if the world
guarantees that the memory at address $a$ will contain a read capability for
address $a$? The definition then just says that a read capability for address
$a$ is safe iff the same read capability for address $a$ is safe. This form of
cyclic reasoning is related to similar problems in languages with recursive
types or higher-order ML-style references, and a standard solution is
step-indexing. Essentially, the cycle is broken by defining safety up to a
certain number of interaction steps. All words would be automatically considered
safe up to $0$ steps and, for example, a read capability would be safe up to $n$
steps if the world guarantees that the words at the corresponding addresses are
safe up to $n - 1$ steps. We can then prove that the above read capability for
address $a$ is safe up to any number of steps.

\paragraph{Future worlds}
So worlds are defined as a set of invariants on the memory, but what if we
allocate fresh memory through malloc? We may want to establish new invariants
for this freshly allocated memory and be sure that the adversary will also
respect those (if we don't provide them with capabilities through which the new
invariants can be broken). To accomodate this, we allow worlds to evolve, for
example by adding additional invariants for freshly allocated memory. Formally,
we define valid ways for a world $W$ to evolve into a new world $W'$ through the
future-world relation $W' \future W$ and we ensure that the set of safe words in
world $W$ must remain safe in any future world $W'$. Defining safety w.r.t. a
notion of evolvable worlds makes our logical relation into a \emph{Kripke}
logical relation.

\paragraph{Invariants and Recursive Worlds}
So worlds group a set of memory invariants, but how are they actually defined
formally? We represent each invariant by a region $r \in \Regions$. We will see
later that regions contain state machines to support a notion of evolvable
invariants, but in every state, they also contain a predicate $H$ that defines
which memory segments are acceptable in the current state of the invariant.
Unfortunately, it is not enough to just take $H \in \powerset{\MemSegments}$,
because sometimes the invariant may itself be world-dependent. For example, we
may want to express invariants like ``the memory at address $50$ contains a
value that is safe in the current world''. As explained, worlds may evolve, and
the set of safe values may grow in future worlds, so then we need to take $H \in
\Worlds \rightarrow \powerset{\MemSegments}$. We then end up with worlds
containing regions and regions that are world-indexed, i.e. the set of worlds is
recursively defined. Nevertheless, this problem is not essential: we will see
how such a recursive definition can again be accomodated using techniques from
the literature (essentially an advanced application of step-indexing).

\paragraph{Local capabilities}
A final thing we explain here is how our results take into account local
capabilities and their special treatment by the hardware. Our idea here is to
introduce a distinction between temporary and permanent invariants. At any point
in a proof, we may decide to revoke temporary invariants. Global capabilities
are then guaranteed to remain safe, but local ones aren't. Concretely, we will
perform such revocations when we need to break the validity of capabilities that
we previously provided to the adversary as a local capability (for example,
return or stack pointers from a previous invocation). In such situations, the
hardware guarantees that global capabilities (which cannot have write-local
permission because of a central assumption\dominique{highlight earlier}) cannot
have been used to store local capabilities and therefore remain safe even after
revoking temporary invariants. Formally, this leads to two distinct future world
relations: we say that $W' \futurewk W$ if a future world $W'$ preserves
temporary invariants and $W' \futurestr W$ if it doesn't. Interestingly, this
is a variant of a notion of public/private future worlds that has been previously
used in the literature (see Section~\ref{sec:discussion}).
%

\subsection{Worlds}
\begin{figure}
  \centering
  \includegraphics{w11}
  \caption{An illustration of local/global capabilities and temporary/permanent
    regions. The table represents a memory and the colored fields are regions
    governing a memory range. Global capabilities cannot depend on temporary
    regions.}
  \label{fig:cap-world}
\end{figure}
% \dominique{Perhaps we could first try to convey the intuition of a permanent
%   world with a temporary overlay, with global capabilities ignoring the overlay.
%   Perhaps we could add a depiction of the world to Figure 5?}
%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}
Formally, a world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, to describe a protocol for the region's memory,
similarly to what has been used in logical relations for high-level
languages~\citep{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. Protocols
imposed by permanent regions stay in place indefinitely. Any capability, local
or global, can depend on these protocols. Protocols imposed by temporary regions
can be revoked at certain points during execution, on the condition that we
clear local capabilities when this happens. This means that local capabilities
can safely depend on the protocols imposed by temporary regions, but global
capabilities cannot, since a global capability may outlive a temporary region
that is revoked. This is illustrated in \figurename~\ref{fig:cap-world}.
